Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-06-01 appendix.tex
changed argument to identifiability

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

@@

\subsection*{Form of the likelihood}

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
We show the Hessian of the joint likelihood term is negative definite, which implies there is a unique maximizer of \eqref{eq:profile_likelihood}.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,2}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,3}(x_2)\\
&\qquad +\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,4}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,5}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
We have the second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
As $x_1,y_1,x_2,y_2\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood is negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
    \hline
     $\{1\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
                \end{array}$\\
    \hline
     $\{1,2\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
                \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
All values multiplied by $1/32$.}
\label{tab:edge-cases}
\end{table}

We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.

The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
\[
(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
\]
or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.

Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

@@

\subsection*{Identifiability criteria and form of the likelihood}

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
In order for our maximum likelihood problem to be well-posed, we require identifiability conditions on the branch parameters.
We show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference yields an identifiable maximum likelihood inference procedure.

We first show that the existence of a unique maximizer implies $x_1=x_2$ and $y_1=y_2$.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood then $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.

It remains to show the converse: if $x_1=x_2$ and $y_1=y_2$ then we obtain a unique estimate of $\tilde{t}$ when performing joint inference maximizing \eqref{eq:profile_likelihood}.
To do so, we show the Hessian of the joint likelihood term is negative definite.
First, we decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Assume $x_1=x_2=x$ and $y_1=y_2=y$ so that $t=\{x,y,x,y,w\}$.
We restrict our attention to $x,y\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
where data is generated on the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=x^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and thus the Hessian of $\shannonDivergence$ is negative definite.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, t).
\]
Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a single partial likelihood function when maximizing over ancestral state splits.
Then, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\tilde{\ell}_{\tau_1,t^*}(\tau_1, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y}(y)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x}(x).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x$ and the initial $1/32$ constant, is
\[
h_{j,x}(x) \propto (1+x)^{e_j}(1-x)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x). \label{eq:partial-x-form}
\end{align}
We have the second derivative
\[
\tilde{\ell}''(x) = -\left(\frac{\sum_j c_je_j}{(1+x)^2}+\frac{1-\sum_j c_je_j}{(1-x)^2}\right).
\]
As $x,y\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood is negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.

We now consider the case where there may be more than one functional form that maximizes the partial likelihood over ancestral state splits for a site split.
Table~\ref{tab:likelihoods} shows that an ambiguity in the form of the partial likelihood occurs in one of four cases:
\[
1: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 > (1+x)^2(1+w)(1-y)^2,
\]
\[
2: (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1-x)^2(1+w)(1+y)^2,
\]
\[
3: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2 > (1+x)^2(1-w)(1+y)^2,
\]
or
\[
4: (1-x)^2(1+w)(1+y)^2 = (1+x)^2(1-w)(1+y)^2 = (1+x)^2(1+w)(1-y)^2.
\]
In case 1, the equality implies $w=2x/(1+x^2)$.
Plugging this into Table~\ref{tab:likelihoods}, we see the partial log-likelihood, again suppressing constants with respect to $x$ and the initial $1/32$ constant, is
\begin{align*}
\tilde{\ell}(x) &\propto (2-p_1-p_3-p_{12}-p_{23}-2p_{13})\log(1+x) \\
&\qquad + (p_1+p_3+p_{12}+p_{23}+2p_{13})\log(1-x) \\
&\qquad + 2\log(1+x)-\log(1+x^2).
\end{align*}
Call
\[
c := p_1+p_3+p_{12}+p_{23}+2p_{13}
\]
so that straightforward calculus reveals
\begin{align*}
\tilde{\ell}''(x) &= \frac{4x^2}{(1+x^2)^2}-\frac{4-c}{(1+x)^2}-\frac{c}{(1-x)^2}\\
                  &= \left(\frac{4x^2}{(1+x^2)^2}-\frac{4}{(1+x)^2}\right)+c\left(\frac{1}{(1+x)^2}-\frac{1}{(1-x)^2}\right).
\end{align*}
The term on the right is negative as $c \ge 0$ and $0 < x < 1$.
The term on the left can be written as
\[
\frac{4}{(\frac{1}{x}+x)^2}-\frac{4}{(1+x)^2},
\]
which for $0<x<1$ is negative so that $\tilde{\ell}''(x) \le 0$.
Case 2 follows the same argument but for $y$.
In case 3, the ambiguity implies $x=y$, which yields a partial log-likelihood of the form \eqref{eq:partial-x-form} multiplied by two, and thus has negative second derivative.
Finally, case 4 implies $x=y$ and $w=2x/(1+x^2)$, and a combination of the above arguments shows we again have a negative second derivative.
Thus, for $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, there is a unique maximum of $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$ if and only if $x_1=x_2$ and $y_1=y_2$, yielding our identifiability criteria.

Given the above, with the symmetry of the likelihood we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
